Computational methods in landscape ecology

### 2024-09-02 Manuscript.qmd

---
title: "Computational methods in landscape ecology"
keywords: keyword1, keyword2
author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
    attributes:
      - corresponding: boolean
  - name: Jakub Nowosad
    orcid: 0000-0002-1057-3721
    email: nowosad.jakub@gmail.com
    affiliations:
      - name: Adam Mickiewicz University, Poznan, Poland
      - name: University of Münster, Münster, Germany
  - name: Marti Bosch
    orcid: 0000-0001-8735-9144
    email: marti.bosch@epfl.ch
    affiliations:
      - name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
  - name: Martin Jung
    orcid: 0000-0002-7569-1390
    email: jung@iiasa.ac.at
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
format:
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header: 
      text: | 
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
papersize: a4paper
fontsize: 10pt
geometry: margin=25mm
indent: 2m
linestretch: 1.5
bibliography: CLER.bib
csl: springer-basic-author-date.csl
---

@@

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/03-connectivity.qmd >}}
{{< include Sections/05-simulations.qmd >}}
{{< include Sections/08-software.qmd >}}

>>>>

---
title: "Computational methods in landscape ecology"
keywords: keyword1, keyword2
author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
    attributes:
      - corresponding: boolean
  - name: Jakub Nowosad
    orcid: 0000-0002-1057-3721
    email: nowosad.jakub@gmail.com
    affiliations:
      - name: Adam Mickiewicz University, Poznan, Poland
      - name: University of Münster, Münster, Germany
  - name: Marti Bosch
    orcid: 0000-0001-8735-9144
    email: marti.bosch@epfl.ch
    affiliations:
      - name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
  - name: Martin Jung
    orcid: 0000-0002-7569-1390
    email: jung@iiasa.ac.at
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
format:
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header: 
      text: | 
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
papersize: a4paper
fontsize: 10pt
geometry: margin=25mm
indent: 2m
linestretch: 1.5
number-sections: true
bibliography: CLER.bib
csl: springer-basic-author-date.csl
---

@@

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/03-connectivity.qmd >}}
{{< include Sections/05-simulations.qmd >}}
{{< include Sections/07-spatial-planning.qmd >}}
{{< include Sections/08-software.qmd >}}

### 2024-09-02 Manuscript.tex

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\usepackage[margin=25mm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\section{Spatial patterns}\label{spatial-patterns}

@@

\section{Connectivity}\label{connectivity}

@@

\section{Simulation models}\label{simulation-models}

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\usepackage[margin=25mm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\section{Spatial patterns}\label{sec-spatialpattern}

@@

\section{Connectivity}\label{sec-connectivity}

@@

\section{Simulation models}\label{sec-simulation}

@@

\section{Spatial conservation planning}\label{sec-spatialplanning}

Spatial planning broadly considers the use of decision theory to
identify and allocate areas to specific purposes, such as reserve
selection through spatial conservation prioritization approaches
(Margules and Pressey 2000; Kukkala and Moilanen 2013; Jung et al.
2024). In the context of landscape-ecology it provides the critical link
between the measuring of patterns and processes (see section
Section~\ref{sec-spatialpattern}), and outcomes related to actual
decision making by distinguishing between structure, function and scales
(Hersperger et al. 2021). Although - as integrative approach - most
spatial planning approaches can benefit from other computational
advances in landscape ecology, such as for example through using
simulation model outputs (see section Section~\ref{sec-simulation}) as
input features to the planning (see also Daigle et al. (2020)), there
have also been several recent frontiers in directly considering
landscape patterns directly in spatial planning.

Spatial patterns in terms of shape and patterns have always been of
interest to spatial planning. Generally, such patterns influence
planning outcomes because of their spatial \emph{compactness},
\emph{contiguity} and \emph{connectedness}. Aggregation can be important
because of ecological reasoning or practical feasibility, for example by
ensuring that any selected areas satisfy certain minimum patch size
constraints (Smith et al. 2010; Justeau-Allaire et al. 2021). Novel
computational approaches make use of graph theory to identify those
areas in a landscape more likely to be of values as compact core habitat
and reducing boundary exposure (Weerasena et al. 2023). Similarly
contiguity and connectivity are often seen as important to ensure that
landscapes and patches of specific characteristics are connected, such
as for example river networks (Beger et al. 2010). In particular the
direct consideration of connectivity in spatial conservation planning
has received much recent attention and computational methods are
increasingly developed (Hanson et al. 2019a, 2022; Daigle et al. 2020;
Beger et al. 2022).

Particular noteworthy here are two advances in incorporating landscape
indices in spatial planning. One is the use of graph theory in linear
programming to guide conservation and restoration efforts
(Justeau-Allaire et al. 2021; Hamonic et al. 2023). For example, in a
New Caledonia case study Justeau-Allaire et al. (2021) showed how theory
and indicators from landscape ecology to determine where restoration
efforts would best be located to maximize broader landscape connectivity
(Justeau-Allaire et al. 2021).

A second main advance is the increasing use of deep reinforcement
learning (RL), a machine learning approach based not on patterns but
agents and pathway, in spatial planning. RL can be used to identify
cost-efficient solutions to area-based conservation planning (Silvestro
et al. 2022) or typical connectivity problems (Equihua et al. 2024),
where the objective is to identify best possible improvements to
landscape connectivity indices such as the Integral Index of
Connectivity. Advantages of RL are that they are able to easier
incorporate linear and non-linear functions and spatial-temporal
processes in the planning and are generally more readibly scalable to
larger spatial extent (Equihua et al. 2024), which however comes with
the downside of decreasing interpretability.

\subsection{Open-source software in spatial conservation
planning}\label{sec-spatialplanning-software}

Multiple open-source software solutions have been developed to support
further integration of landscape ecology theory and indicators into
spatial planning. Most of these software are based on linear or
constrained programming approaches, which however require a solver to
create any outputs. Unfortunately there is still a gap between
state-of-the art open-source and proprietary solvers in terms of
computational efficiency (Hanson et al. 2019b). Noteworthy software is
the \emph{prioritizr} R-package (Hanson et al. 2024) which supports a
range of compactness and contiguity constraints and penalties, and
generally allows for the flexibly creation of planning problems in
multiple configurations. Similarly, the \emph{coco} package makes use of
connectivity constraints to identify potential outcomes that minimize
further fragmentation of landscapes (Mantgem et al. 2023). The new
\emph{restoptr} R-package makes use of constrained programming for those
areas to be restored in a region that would most effectively contribute
to greater landscape connectivity (Justeau-Allaire et al. 2023).
Innovatively, it uses landscape ecology indices not as constraints, but
as specific targets (e.g.~where one could best maximize the effective
mesh size in a region). The CAPTAIN software suite enables the
application of reinforcement learning for area-based conservation
efforts (Silvestro et al. 2022). Overall, most of these software however
still require substantial prior knowledge on (spatial) optimization and
statistics, which can be an entry barrier to more widespread uptake.

@@

\bibitem[\citeproctext]{ref-beger2010incorporating}
Beger M, Linke S, Watts M, et al (2010) Incorporating asymmetric
connectivity into spatial decision making for conservation. Conservation
Letters 3:359--368

\bibitem[\citeproctext]{ref-beger2022demystifying}
Beger M, Metaxas A, Balbar AC, et al (2022) Demystifying ecological
connectivity for actionable spatial conservation planning. Trends in
ecology \& evolution 37:1079--1091

@@

\bibitem[\citeproctext]{ref-daigle2020operationalizing}
Daigle RM, Metaxas A, Balbar AC, et al (2020) Operationalizing
ecological connectivity in spatial conservation planning with marxan
connect. Methods in Ecology and Evolution 11:570--579

@@

\bibitem[\citeproctext]{ref-equihua2024connectivity}
Equihua J, Beckmann M, Seppelt R (2024) Connectivity conservation
planning through deep reinforcement learning. Methods in Ecology and
Evolution 15:779--790

@@

\bibitem[\citeproctext]{ref-hamonic2023cumulative}
Hamonic F, Couëtoux B, Vaxès Y, Albert CH (2023) Cumulative effects on
habitat networks: How greedy should we be? Biological Conservation
282:110066

@@

\bibitem[\citeproctext]{ref-hanson2019conventional}
Hanson JO, Fuller RA, Rhodes JR (2019a) Conventional methods for
enhancing connectivity in conservation planning do not always maintain
gene flow. Journal of Applied Ecology 56:913--922

\bibitem[\citeproctext]{ref-hanson2024prioritizr}
Hanson JO, Schuster R, Morrell N, et al (2024)
\href{https://CRAN.R-project.org/package=prioritizr}{Prioritizr:
Systematic conservation prioritization in r}

\bibitem[\citeproctext]{ref-hanson2019optimality}
Hanson JO, Schuster R, Strimas-Mackey M, Bennett JR (2019b) Optimality
in prioritizing conservation projects. Methods in Ecology and Evolution
10:1655--1663

\bibitem[\citeproctext]{ref-hanson2022comparison}
Hanson JO, Vincent J, Schuster R, et al (2022) A comparison of
approaches for including connectivity in systematic conservation
planning. Journal of Applied Ecology 59:2507--2519

\bibitem[\citeproctext]{ref-hersperger2021landscape}
Hersperger AM, Grădinaru SR, Pierri Daunt AB, et al (2021) Landscape
ecological concepts in planning: Review of recent developments.
Landscape ecology 1--17

@@

\bibitem[\citeproctext]{ref-jung2024assessment}
Jung M, Alagador D, Chapman M, et al (2024) An assessment of the state
of conservation planning in europe. Philosophical Transactions of the
Royal Society B 379:20230015

@@

\bibitem[\citeproctext]{ref-justeau2023restoptr}
Justeau-Allaire D, Hanson JO, Lannuzel G, et al (2023) Restoptr: An r
package for ecological restoration planning. Restoration Ecology
31:e13910

\bibitem[\citeproctext]{ref-justeau2021constrained}
Justeau-Allaire D, Vieilledent G, Rinck N, et al (2021) Constrained
optimization of landscape indices in conservation planning to support
ecological restoration in new caledonia. Journal of Applied Ecology
58:744--754

@@

\bibitem[\citeproctext]{ref-kukkala2013core}
Kukkala AS, Moilanen A (2013) Core concepts of spatial prioritisation in
systematic conservation planning. Biological Reviews 88:443--464

@@

\bibitem[\citeproctext]{ref-van2023coco}
Mantgem ES van, Hillebrand J, Rose L, Klau GW (2023) Coco: Conservation
design for optimal ecological connectivity. Frontiers in Ecology and
Evolution 11:1149571

\bibitem[\citeproctext]{ref-margules2000systematic}
Margules CR, Pressey RL (2000) Systematic conservation planning. Nature
405:243--253

@@

\bibitem[\citeproctext]{ref-silvestro2022improving}
Silvestro D, Goria S, Sterner T, Antonelli A (2022) Improving
biodiversity protection through artificial intelligence. Nature
sustainability 5:415--424

@@

\bibitem[\citeproctext]{ref-smith2010approach}
Smith RJ, Di Minin E, Linke S, et al (2010) An approach for ensuring
minimum protected area size in systematic conservation planning.
Biological Conservation 143:2525--2531

@@

\bibitem[\citeproctext]{ref-weerasena2023sequential}
Weerasena L, Shier D, Tonkyn D, et al (2023) A sequential approach to
reserve design with compactness and contiguity considerations.
Ecological Modelling 478:110281

### 2024-09-02 Sections/07-spatial-planning.qmd
Updating software section and tiny cosmetic changes

# Spatial planning

Landscape ecology is fun [@Turner1989].
There are many things to explore [@With2019].

>>>>

# Spatial conservation planning {#sec-spatialplanning}

Spatial planning broadly considers the use of decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization approaches [@margules2000systematic; @kukkala2013core; @jung2024assessment]. In the context of landscape-ecology it provides the critical link between the measuring of patterns and processes (see section @sec-spatialpattern), and outcomes related to actual decision making by distinguishing between structure, function and scales [@hersperger2021landscape]. Although - as integrative approach - most spatial planning approaches can benefit from other computational advances in landscape ecology, such as for example through using simulation model outputs (see section @sec-simulation) as input features to the planning (see also @daigle2020operationalizing), there have also been several recent frontiers in directly considering landscape patterns directly in spatial planning.

Spatial patterns in terms of shape and patterns have always been of interest to spatial planning. Generally, such patterns influence planning outcomes because of their spatial *compactness*, *contiguity* and *connectedness*. Aggregation can be important because of ecological reasoning or practical feasibility, for example by ensuring that any selected areas satisfy certain minimum patch size constraints [@smith2010approach;@justeau2021constrained]. Novel computational approaches make use of graph theory to identify those areas in a landscape more likely to be of values as compact core habitat and reducing boundary exposure [@weerasena2023sequential]. Similarly contiguity and connectivity are often seen as important to ensure that landscapes and patches of specific characteristics are connected, such as for example river networks [@beger2010incorporating]. In particular the direct consideration of connectivity in spatial conservation planning has received much recent attention and computational methods are increasingly developed [@hanson2019conventional; @daigle2020operationalizing; @hanson2022comparison; @beger2022demystifying].

Particular noteworthy here are two advances in incorporating landscape indices in spatial planning. One is the use of graph theory in linear programming to guide conservation and restoration efforts [@justeau2021constrained; @hamonic2023cumulative]. For example, in a New Caledonia case study @justeau2021constrained showed how theory and indicators from landscape ecology to determine where restoration efforts would best be located to maximize broader landscape connectivity. The direct consideration of graph theoretic approaches in spatial networks has the potential to be provide more cost-effective and precise solutions to the design of reserve networks [@hamonic2023cumulative].

A second main advance is the increasing use of deep reinforcement learning (RL), a machine learning approach based not on patterns but agents and pathway, in spatial planning. RL can be used to identify cost-efficient solutions to area-based conservation planning [@silvestro2022improving] or typical connectivity problems [@equihua2024connectivity], where the objective is to identify best possible improvements to landscape connectivity indices such as the Integral Index of Connectivity. Advantages of RL are that they are able to easier incorporate linear and non-linear functions and spatial-temporal processes in the planning and are generally more readibly scalable to larger spatial extent [@equihua2024connectivity], which however comes with the downside of decreasing interpretability.

## Open-source software in spatial conservation planning {#sec-spatialplanning-software}

Multiple open-source software solutions have been developed to support further integration of landscape ecology theory and indicators into spatial planning. Most of these software are based on linear or constrained programming approaches, which however require a solver to create any outputs. Unfortunately there is still a gap between state-of-the art open-source and proprietary solvers in terms of computational efficiency [@hanson2019optimality].

Noteworthy software is the *prioritizr* R-package [@hanson2024prioritizr] which supports a range of compactness and contiguity constraints and penalties, and generally allows for the flexibly creation of planning problems in multiple configurations. 
Similarly, the *coco* package makes use of connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of landscapes [@van2023coco]. The new *restoptr* R-package makes use of constrained programming for those areas to be restored in a region that would most effectively contribute to greater landscape connectivity [@justeau2023restoptr]. Innovatively, it uses landscape ecology indices not as constraints, but as specific targets (e.g. where one could best maximize the effective mesh size in a region). The *CAPTAIN* software suite enables the application of reinforcement learning for area-based conservation efforts [@silvestro2022improving].
Overall, most of these software however still require substantial prior knowledge on (spatial) optimization and statistics, which can be an entry barrier to more widespread uptake.
